---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Set Global Options

```{r setoptions, echo=TRUE}
#load libs
library(knitr)
library(dplyr)
library(stringr)
library(lattice)

#set echo on
opts_chunk$set(echo = TRUE)

```


## Loading and preprocessing the data


```{r loaddata}  
#create dir if not exist
if (!file.exists("datadir"))
{
  dir.create("datadir")
}

#unzip zip into datadir directory
unzip("activity.zip",exdir="./datadir")

#read csv
dataset <- read.csv("./datadir/activity.csv", sep = ",", stringsAsFactors = FALSE, header= TRUE)

#convert to date
dataset$date <- as.Date(dataset$date,"%Y-%m-%d")
```

## What is mean total number of steps taken per day?
```{r sumsteps}
#filter out NA's
datasetMinusNA <- subset(dataset,!is.na(dataset$steps))

#group by date and sum steps
totalStepsPerDay <- summarise(group_by(datasetMinusNA,date),sum(steps))

#plot histogram
hist(totalStepsPerDay$`sum(steps)`,
    main="Histogram of the total number of steps taken each day",xlab="Total Steps")
```

```{r meanAndmedian}
#calculate mean and median
totalStepsPerDayMean <- mean(totalStepsPerDay$`sum(steps)`)
totalStepsPerDayMedian <- median(totalStepsPerDay$`sum(steps)`)
```

For the total number of steps taken per day the mean is `r totalStepsPerDayMean` and the median is `r totalStepsPerDayMedian` 

````{r dumpDataMeanMedian, results="asis"}
#group by date and create mean,median and sum's
stepsPerDayMeanMedian <- summarise(group_by(datasetMinusNA,date),mean(steps),median(steps),sum(steps))

#output raw data for review
kable(stepsPerDayMeanMedian, format = "html")
````


## What is the average daily activity pattern?

```{r avgdaily}
#Requirement:Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

#Take the non-NA dataset
#group by interval and take the average of steps
averageStepsPerInterval <- summarise(group_by(datasetMinusNA,interval),mean(steps))

#pad with zeros the interval then convert to time datatype before plotting
with(averageStepsPerInterval,plot(strptime(str_pad(interval,4,side="left",pad="0"),"%H%M"),
                                   `mean(steps)`,
                                    type = "l",
                                    main="Average daily activity pattern",
                                    xlab="Hours",
                                    ylab="Average number of steps"))
```

```{r calcMaxInterval}
# find max steps
# subset based on max steps and return column 1
# format accordingly
maxInterval <- format(strptime(str_pad(averageStepsPerInterval[which.max(averageStepsPerInterval$`mean(steps)`),1],4,side="left",pad="0"),"%H%M"),"%H:%M")

```


The `r maxInterval` 5-minute interval, on average across all the days in the dataset, contained the maximum number of steps.

## Imputing missing values

```{r missingValues}
totalMissingValues <- sum(is.na(dataset$steps))
```

Total number of missing values in the dataset (i.e. the total number of rows with NAs) = `r totalMissingValues`

I'm using 'step mean' values for specific intervals across all days to populate missing values

```{r fillInValues}
#copy dataset
datasetWithNAsFilledIn <- dataset

#loop over dataframe rows
for (i in 1:nrow(datasetWithNAsFilledIn) )
{
  #extract row
  datasetRow <- datasetWithNAsFilledIn[i,]

  #if NA then
  if (is.na(datasetRow$steps))
  {
    #copy step (column 2) from the average dataframe where interval equals the interval of the current row
    datasetWithNAsFilledIn[i,1] <- subset(averageStepsPerInterval,interval==datasetRow$interval,2)
  }
}

#group by date and sum steps
totalStepsPerDayFilledIN <- summarise(group_by(datasetWithNAsFilledIn,date),sum(steps))

#plot histogram
hist(totalStepsPerDayFilledIN$`sum(steps)`,
    main="Histogram of the total number of steps taken each day for filled in NAs",xlab="Total Steps")

```

```{r meanAndmedianFilledIn}
#calculate mean and median
totalStepsPerDayMeanFilled <- mean(totalStepsPerDayFilledIN$`sum(steps)`)
totalStepsPerDayMedianFilled <- median(totalStepsPerDayFilledIN$`sum(steps)`)
```


For the total number of steps taken per day the mean is `r totalStepsPerDayMeanFilled` and the median is `r totalStepsPerDayMedianFilled` 

````{r dumpDataMeanMedianFilled, results="asis"}
#output raw data for review
stepsPerDayMeanMedianFilled <- summarise(group_by(datasetWithNAsFilledIn,date),mean(steps),median(steps),sum(steps))
kable(stepsPerDayMeanMedianFilled, format = "html")
````


You can see by adding in missing values thatthe frequency of particular ranges has increased. However when you look at the RAW data the filled in values look like anomalies. 

## Are there differences in activity patterns between weekdays and weekends?

```{r daysAndWeekends}
#copy data with NA's filled in
datasetWithFactors <- datasetWithNAsFilledIn

#if weekday is Saturday OR Sunday then set new column to weekend else weekday
datasetWithFactors$weektype <- ifelse(weekdays(datasetWithNAsFilledIn$date)=="Sunday" | weekdays(datasetWithNAsFilledIn$date)=="Saturday" , "weekend", "weekday")

#set as factor datatype
datasetWithFactors$weektype <- as.factor(datasetWithFactors$weektype)


#Take the NA filled In dataset
#group by interval and take the average of steps
averageStepsPerIntervalFilled <- summarise(group_by(datasetWithFactors,weektype,interval),mean(steps))

#pad with zeros the interval then convert to time datatype before plotting
xyplot(`mean(steps)`~interval|weektype, data=averageStepsPerIntervalFilled, 
                                    type="l",
                                    main="Average daily activity pattern",
                                    xlab="Hours",
                                    ylab="Average number of steps",
                                    layout=c(1,2))

```
